VitalLens: Take A Vital Selfie

### 2023-12-11 paper.tex
Changes based on Pieter's feedback; Version submitted to arXiv

\usepackage[final]{assets/techreport}
%\usepackage[nonatbib]{assets/techreport}

@@

\csvreader[before reading=\def\totalparticipantsvvghana{0}\def\totalchunksvvghana{0}\def\totaltimevvghana{0}]{data/vv_ghana_small_summary.csv}{participants=\participants,chunks=\chunks,time=\time}{%
	\pgfmathsetmacro{\totalparticipantsvvghana}{\totalparticipantsvvghana+\participants}%
	\pgfmathsetmacro{\totalchunksvvghana}{\totalchunksvvghana+\chunks}%
	\pgfmathsetmacro{\totaltimevvghana}{\totaltimevvghana+\time}%
}

@@

Video of the human face and upper body has proven to be a rich source of information about an individual's physiological state \cite{verkruysse2008remote}.
In particular, signals embedded in these videos can be harnessed to estimate vital signs such as heart rate and respiratory rate through a process known as Remote Photoplethysmography (rPPG).
This capability holds immense potential for non-invasive and real-time health monitoring applications.

Various rPPG approaches have been proposed, including handcrafted algorithms and ones learned from empirical data.
Handcrafted algorithms offer advantages of fast inference speed and no training data required, but usually lack in accuracy; notable approaches include the original approach G \cite{verkruysse2008remote}, and the more sophisticated CHROM \cite{de2013robust} and POS \cite{wang2017algorithmic}.
Given enough high-quality data to learn from, learning-based approaches such as DeepPhys \cite{chen2018deep} and MTTS-CAN \cite{liu2020multi} hold the promise of greater accuracy, which they have to trade off against inference speed.
There are also other learning-based approaches achieving higher accuracy, which we do not cover here as they are not aiming to achieve real-time inference.

@@

\begin{itemize}
	\item \textbf{Real-time rPPG application.} We introduce VitalLens -- the first widely distributed rPPG application for real time estimation of vitals such as of heart rate and respiratory rate. 
	\item \textbf{Comprehensive evaluation on a large dataset.} We evaluate the performance of VitalLens on multiple datasets, including the large VV-Medium \cite{toye2023vital} with \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvmedium} participants. VitalLens outperforms G, CHROM, POS, DeepPhys, and MTTS-CAN in terms of estimation accuracy. On VV-Medium, VitalLens achieves absolute errors of 0.71 bpm for heart rate estimation, and 0.76 rpm for respiratory rate estimation, while taking 18 ms to run inference.
	\item \textbf{Analysis of factors impacting performance.} We systematically investigate factors impacting estimation performance. We find that \textit{participant movement} and \textit{variation of participant illuminance} are the main factors impacting performance. Compared to approaches like POS, VitalLens is less impacted by darker skin types when estimating heart rates. 
	\item \textbf{Maintaining end-user privacy}. VitalLens runs inference locally, meaning that video and vitals never leave the device.
\end{itemize}

@@

VitalLens is trained on the \textit{PROSIT} and \textit{Vital Videos Ghana} datasets.
The vast majority of training data comes from PROSIT.
For evaluation, we use the entire \textit{Vital Videos Medium} dataset as well as the test sets of the \textit{PROSIT} and \textit{Vital Videos Ghana} datasets.

@@

\subsection{Vital Videos Ghana}

Vital Videos Ghana is a new dataset from the authors of Vital Videos aiming to address the insufficient share of participants with skin types 5 and 6 usually found in rPPG datasets.
It was collected in Ghana, with the majority of participants having skin type 5 or 6.
We use a small instantiation (``VV-Ghana-Small''), which consists of \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvghana} participants.
Both camera and participants are stationary in this dataset.

@@

We pre-process VV-Ghana-Small using the same steps applied to PROSIT.
As part of this step, we create small chunks, calculate summary vitals, and extract further metadata.
Note that for some chunks, the missing respiratory signal was synthetically created using the same procedure as for VV-Medium.

\begin{table}[h!]
 	\caption{VV-Ghana-Small Dataset Size}
 	\label{tab:vv-ghana-small-summary}
 	\centering
  	\csvreader[
		tabular=llll,
    	table head=\toprule Split & \# Participants & \# Chunks & Time \\ \midrule,
    	table foot=\midrule Total & \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvghana} & \sisetup{round-mode=places,round-precision=0}\num{\totalchunksvvghana} & \sisetup{round-mode=places,round-precision=1}\SI{\totaltimevvghana}{\hour} \\ \bottomrule,
    	late after line=\\,
    	before reading={\catcode`\%=12}, after reading={\catcode`\%=14}
  	]{data/vv_ghana_small_summary.csv}{}%
  {\csvcoli & \num{\csvcolii} & \num{\csvcoliii} & \sisetup{round-mode=places,round-precision=1}\SI{\csvcoliv}{\hour} }
\end{table}

@@

\subsection{Final training dataset: PROSIT + VV-Ghana-Small}
\csvreader[before reading=\def\totalparticipantstraining{0}\def\totalchunkstraining{0}\def\totaltimetraining{0}]{data/training_summary.csv}{participants=\participants,chunks=\chunks,time=\time}{%
		\pgfmathsetmacro{\totalparticipantstraining}{\totalparticipantstraining+\participants}%
		\pgfmathsetmacro{\totalchunkstraining}{\totalchunkstraining+\chunks}%
		\pgfmathsetmacro{\totaltimetraining}{\totaltimetraining+\time}%
	}
We combine the training sets of PROSIT and VV-Ghana-Small for training.
As of this writing, this makes a total of \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantstraining} unique participants, \sisetup{round-mode=places,round-precision=0}\num{\totalchunkstraining} chunks or \sisetup{round-mode=places,round-precision=1}\num{\totaltimetraining} hours of data.

@@

\paragraph{Model Validation.}
While developing the model, we train many different versions of our model.
During training, we continuously monitor the model's performance on the validation sets of PROSIT and VV-Ghana-Small as outlined in Section \ref{sec:datasets}.
Since the training dataset is disjoint from both validation sets in terms of participants, this allows us to calculate validation metrics which measure how well each model has learned to generalize in estimating vital signs from video.
We used the validation metrics to select the final model to be used in VitalLens.

@@

\begin{itemize}
	\item The entire VV-Medium dataset, a large and diverse dataset of \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvmedium} participants introduced in Section \ref{sec:datasets}. It includes participants with varying demographics and environmental conditions.
	\item The test set of the PROSIT dataset. This is important as PROSIT includes camera and participant movement, which is to be expected when VitalLens is used in the wild.
	\item The test set of the VV-Ghana-Small dataset.
\end{itemize}

@@

We evaluated VitalLens on three datasets - VV-Medium, the PROSIT test set, and the VV-Ghana-Small test set.
None of the participants included in these datasets were previously seen by VitalLens.
In addition, we also evaluated several other existing methods to allow benchmarking of VitalLens:
The handcrafted methods G, CHROM, and POS, as well as the learning-based methods DeepPhys and MTTS-CAN.
To allow a fair comparison, we trained the latter two on the same training data as VitalLens.

@@

The PROSIT test set has a lower variety in participants and their demographics, but it includes both participant and camera movement.
Results for the PROSIT test set are reported in Table \ref{tab:results-prosit-test} -- again, VitalLens beats all listed methods for all considered evaluation metrics.
However, we observe that here, the absolute errors achieved by VitalLens for heart rate and respiratory rate are 4.62 bpm and 3.76, respectively -- markedly higher than for VV-Medium.
This is a first hint that factors like participant and camera movement are the main drivers of lower estimation accuracies.

\begin{table}[h!]
 	\caption{Vitals estimation results on VV-Ghana-Small test set}
 	\label{tab:results-vv-ghana-small-test}
 	\centering
 	\begin{threeparttable}
  	\csvreader[
		tabular=lcccccc,
    	table head=\toprule Method & \multicolumn{1}{c}{Heart rate} & \multicolumn{2}{c}{Pulse wave} & \multicolumn{1}{c}{Respiratory rate\tnote{a}} & \multicolumn{2}{c}{Respiration wave\tnote{a}} \\ & MAE $\downarrow$ & SNR $\uparrow$ & $r \uparrow$ & MAE $\downarrow$ & SNR $\uparrow$ & $r \uparrow$ \\ \midrule,
    	table foot=\bottomrule,
    	late after line=\\,
    	before reading={\catcode`\%=12}, after reading={\catcode`\%=14}
  	]{data/results_vv_ghana_test.csv}{}%
  {\csvcoli & \printMetric{\csvcolii} & \printMetric{\csvcoliii} & \printMetric{\csvcoliv} & \printMetric{\csvcolv} & \printMetric{\csvcolvi} & \printMetric{\csvcolvii} }
  \begin{tablenotes}
	\item[a] Missing respiratory signals were synthetically created and manually verified.
	\end{tablenotes}
  \end{threeparttable}
\end{table}

Finally, Table \ref{tab:results-vv-ghana-small-test} lists the results for the VV-Ghana-Small test set.
The results paint a similar picture as the two previous datasets, with VitalLens beating out the other methods.
In comparison to VV-Medium, we observe that the errors are generally higher - in this case, this is caused by the demographic differences between the participants.
We further investigate this in the following sections.

@@

Overall, the results confirm that the behavioral factors of participant movement and variation in the illuminance of the participant's face dominate the results with have a significant negative impact on estimation performance.
It is interesting to note that at least for these datasets, the effect of illuminance variation seems to have a greater impact than participant movement.
Our model performance appears to be slightly impacted by the heart rate itself.
We also confirm previous findings that rPPG models can struggle with darker skin types, but cannot confirm findings that estimation is more accurate for females \cite{nowara2020meta}.
Additionally, we observe that estimation appears to be more accurate for older participants.

@@

\begin{figure}[h!]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        ylabel=SNR,
        height=5cm,
        width=.8\textwidth,
        legend style={
            at={(0.98,.96)},
            anchor=north east,
            legend columns=-1
        },
        ybar,
        bar width=14.,
        xtick=data,
        xticklabels from table={\datatable}{bin},
        xticklabel style={align=center, text width=3cm},
    ]
    \addplot+[
    	red!60,
        ybar,
        error bars/.cd,
        y dir=both,
        y explicit,
    ] table[x expr=\coordindex, y=pulse_snr_mean_vl, y error plus=pulse_snr_sd_vl] {\datatable};
    \addlegendentry{VitalLens};
    \addplot+[
        green!60,
        ybar,
        error bars/.cd,
        y dir=both,
        y explicit,
    ] table[x expr=\coordindex, y=pulse_snr_mean_vl_64, y error plus=pulse_snr_sd_vl_64] {\datatable};
    \addlegendentry{VitalLens*};
    \end{axis}
  \end{tikzpicture}
  \caption{Comparing signal-to-noise ratio between VitalLens and VitalLens* on VV-Medium, grouped by participant skin types. VitalLens* was solely trained on PROSIT, while VitalLens* was trained on both PROSIT and VV-Ghana-Small.}
  \label{fig:impact-skin-type-comparison}
\end{figure}

Next, we want to investigate to what degree the inclusion of VV-Ghana-Small in our training data lessened this skin type effect.
For this purpose, we compare VitalLens with \textit{VitalLens*}, a version of VitalLens that is solely trained on the PROSIT test set.
The training set of PROSIT by itself only includes 2 individuals of skin type 6, a number that is massively increased in the combined training data.
Both VitalLens and VitalLens* are then evaluated on VV-Medium.
Figure \ref{fig:impact-skin-type-comparison} compares the results, again grouped by skin type.
It is evident that the addition of VV-Ghana-Small led to a sizeable improvement mainly for skin types 5 and 6, but also skin types 3 and 4.

@@

\section*{References}

>>>>

\usepackage[final,nonatbib]{assets/techreport}

@@

\csvreader[before reading=\def\totalparticipantsvvafrica{0}\def\totalchunksvvafrica{0}\def\totaltimevvafrica{0}]{data/vv_africa_small_summary.csv}{participants=\participants,chunks=\chunks,time=\time}{%
	\pgfmathsetmacro{\totalparticipantsvvafrica}{\totalparticipantsvvafrica+\participants}%
	\pgfmathsetmacro{\totalchunksvvafrica}{\totalchunksvvafrica+\chunks}%
	\pgfmathsetmacro{\totaltimevvafrica}{\totaltimevvafrica+\time}%
}

@@

Video of the human face and upper body has proven to be a rich source of information about an individual's physiological state \cite{verkruysse2008remote}.
In particular, signals embedded in these videos can be harnessed to estimate vital signs through a process known as Remote Photoplethysmography (rPPG).
This capability holds immense potential for non-invasive and real-time health monitoring applications.

Various rPPG approaches have been proposed, including handcrafted algorithms and ones learned from empirical data.
Handcrafted algorithms offer advantages of fast inference speed and no training data required, but usually lack in accuracy; notable approaches include the original approach G \cite{verkruysse2008remote}, and the more sophisticated CHROM \cite{de2013robust} and POS \cite{wang2017algorithmic}.
Given enough high-quality data to learn from, learning-based approaches such as DeepPhys \cite{chen2018deep} and MTTS-CAN \cite{liu2020multi} hold the promise of greater accuracy, which is traded off against inference speed.
There are also other learning-based approaches, which we do not cover here as they are not aiming to achieve real-time inference.

@@

\begin{itemize}
	\item \textbf{Real-time rPPG application.} We introduce VitalLens -- the first widely distributed rPPG application for real time estimation of vitals such as of heart rate and respiratory rate.\footnote{VitalLens is currently in private beta test and/or under review by Apple. This will be updated with a public App Store link once available.} 
	\item \textbf{Comprehensive evaluation on a large dataset.} We evaluate the performance of VitalLens on multiple datasets, including the large VV-Medium \cite{toye2023vital} with \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvmedium} participants. VitalLens outperforms G, CHROM, POS, DeepPhys, and MTTS-CAN in terms of estimation accuracy. On VV-Medium, VitalLens achieves absolute errors of 0.71 bpm for heart rate estimation, and 0.76 rpm for respiratory rate estimation, while taking 18 ms to run inference.
	\item \textbf{Analysis of factors impacting performance.} We systematically investigate factors impacting estimation performance. We find that \textit{participant movement} and \textit{variation of participant illuminance} are the main factors impacting performance. Compared to approaches like POS, VitalLens is less impacted by darker skin types when estimating heart rates. 
	\item \textbf{Maintaining end-user privacy}. VitalLens runs inference locally, meaning that video and vitals never leave the device.
\end{itemize}

@@

VitalLens is trained on the \textit{PROSIT} and \textit{Vital Videos Africa} datasets.
The vast majority of training data comes from PROSIT.
For evaluation, we use the entire \textit{Vital Videos Medium} dataset as well as the test sets of the \textit{PROSIT} and \textit{Vital Videos Africa} datasets.

@@

\subsection{Vital Videos Africa}

Vital Videos Africa is a new dataset from the authors of Vital Videos aiming to address the insufficient share of participants with skin types 5 and 6 usually found in rPPG datasets.
It was collected in Ghana, with the majority of participants having skin type 5 or 6.
We use a small instantiation (``VV-Africa-Small''), which consists of \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvafrica} participants.
Both camera and participants are stationary in this subset.

@@

We pre-process VV-Africa-Small using the same steps applied to PROSIT.
As part of this step, we create small chunks, calculate summary vitals, and extract further metadata.
Note that for some chunks, the missing respiratory signal was synthetically created using the same procedure as for VV-Medium.

\begin{table}[h!]
 	\caption{VV-Africa-Small Dataset Size}
 	\label{tab:vv-africa-small-summary}
 	\centering
  	\csvreader[
		tabular=llll,
    	table head=\toprule Split & \# Participants & \# Chunks & Time \\ \midrule,
    	table foot=\midrule Total & \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvafrica} & \sisetup{round-mode=places,round-precision=0}\num{\totalchunksvvafrica} & \sisetup{round-mode=places,round-precision=1}\SI{\totaltimevvafrica}{\hour} \\ \bottomrule,
    	late after line=\\,
    	before reading={\catcode`\%=12}, after reading={\catcode`\%=14}
  	]{data/vv_africa_small_summary.csv}{}%
  {\csvcoli & \num{\csvcolii} & \num{\csvcoliii} & \sisetup{round-mode=places,round-precision=1}\SI{\csvcoliv}{\hour} }
\end{table}

@@

\subsection{Final training dataset: PROSIT + VV-Africa-Small}
\csvreader[before reading=\def\totalparticipantstraining{0}\def\totalchunkstraining{0}\def\totaltimetraining{0}]{data/training_summary.csv}{participants=\participants,chunks=\chunks,time=\time}{%
		\pgfmathsetmacro{\totalparticipantstraining}{\totalparticipantstraining+\participants}%
		\pgfmathsetmacro{\totalchunkstraining}{\totalchunkstraining+\chunks}%
		\pgfmathsetmacro{\totaltimetraining}{\totaltimetraining+\time}%
	}
We combine the training sets of PROSIT and VV-Africa-Small for training.
As of this writing, this makes a total of \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantstraining} unique participants, \sisetup{round-mode=places,round-precision=0}\num{\totalchunkstraining} chunks or \sisetup{round-mode=places,round-precision=1}\num{\totaltimetraining} hours of data.

@@

\paragraph{Model Validation.}
While developing the model, we train many different versions of our model.
During training, we continuously monitor the model's performance on the validation sets of PROSIT and VV-Africa-Small as outlined in Section \ref{sec:datasets}.
Since the training dataset is disjoint from both validation sets in terms of participants, this allows us to calculate validation metrics which measure how well each model has learned to generalize in estimating vital signs from video.
We used the validation metrics to select the final model to be used in VitalLens.

@@

\begin{itemize}
	\item The entire VV-Medium dataset, a large and diverse dataset of \sisetup{round-mode=places,round-precision=0}\num{\totalparticipantsvvmedium} participants introduced in Section \ref{sec:datasets}. It includes participants with varying demographics and environmental conditions.
	\item The test set of the PROSIT dataset. This is important as PROSIT includes camera and participant movement, which is to be expected when VitalLens is used in the wild.
	\item The test set of the VV-Africa-Small dataset.
\end{itemize}

@@

We evaluated VitalLens on three datasets - VV-Medium, the PROSIT test set, and the VV-Africa-Small test set.
None of the participants included in these datasets were previously seen by VitalLens.
In addition, we also evaluated several other existing methods to allow benchmarking of VitalLens:
The handcrafted methods G, CHROM, and POS, as well as the learning-based methods DeepPhys and MTTS-CAN.
To allow a fair comparison, we trained the latter two on the same training data as VitalLens.

@@

The PROSIT test set has a lower variety in participants and their demographics, but it includes both participant and camera movement.
Results for the PROSIT test set are reported in Table \ref{tab:results-prosit-test} -- again, VitalLens beats all listed methods for all considered evaluation metrics.
However, we observe that here, the absolute errors achieved by VitalLens for heart rate and respiratory rate are 4.62 bpm and 3.76 bpm, respectively -- markedly higher than for VV-Medium.
This is a first hint that factors like participant and camera movement are the main drivers of lower estimation accuracies.

\begin{table}[h!]
 	\caption{Vitals estimation results on VV-Africa-Small test set}
 	\label{tab:results-vv-africa-small-test}
 	\centering
 	\begin{threeparttable}
  	\csvreader[
		tabular=lcccccc,
    	table head=\toprule Method & \multicolumn{1}{c}{Heart rate} & \multicolumn{2}{c}{Pulse wave} & \multicolumn{1}{c}{Respiratory rate\tnote{a}} & \multicolumn{2}{c}{Respiration wave\tnote{a}} \\ & MAE $\downarrow$ & SNR $\uparrow$ & $r \uparrow$ & MAE $\downarrow$ & SNR $\uparrow$ & $r \uparrow$ \\ \midrule,
    	table foot=\bottomrule,
    	late after line=\\,
    	before reading={\catcode`\%=12}, after reading={\catcode`\%=14}
  	]{data/results_vv_africa_test.csv}{}%
  {\csvcoli & \printMetric{\csvcolii} & \printMetric{\csvcoliii} & \printMetric{\csvcoliv} & \printMetric{\csvcolv} & \printMetric{\csvcolvi} & \printMetric{\csvcolvii} }
  \begin{tablenotes}
	\item[a] Missing respiratory signals were synthetically created and manually verified.
	\end{tablenotes}
  \end{threeparttable}
\end{table}

Finally, Table \ref{tab:results-vv-africa-small-test} lists the results for the VV-Africa-Small test set.
The results paint a similar picture as the two previous datasets, with VitalLens beating out the other methods.
In comparison to VV-Medium, we observe that the errors are generally higher - in this case, this may be caused by the demographic differences between the participants and more challenging lighting conditions.
We further investigate this in the following sections.

@@

Overall, the results confirm that the behavioral factors of participant movement and variation in the illuminance of the participant's face dominate the results with a significant negative impact on estimation performance.
It is interesting to note that at least for these datasets, the effect of illuminance variation seems to have a greater impact than participant movement.
Our model performance appears to be slightly impacted by the heart rate itself.
We also confirm previous findings that rPPG models can struggle with darker skin types, but cannot confirm findings that estimation is more accurate for females \cite{nowara2020meta}.
Additionally, we observe that estimation appears to be more accurate for older participants.

@@

\begin{figure}[h!]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        ylabel=SNR,
        height=5cm,
        width=.8\textwidth,
        legend style={
            at={(0.98,.96)},
            anchor=north east,
            legend columns=-1
        },
        ybar,
        bar width=14.,
        xtick=data,
        xticklabels from table={\datatable}{bin},
        xticklabel style={align=center, text width=3cm},
    ]
    \addplot+[
    	red!60,
        ybar,
        error bars/.cd,
        y dir=both,
        y explicit,
    ] table[x expr=\coordindex, y=pulse_snr_mean_vl, y error plus=pulse_snr_sd_vl] {\datatable};
    \addlegendentry{VitalLens};
    \addplot+[
        green!60,
        ybar,
        error bars/.cd,
        y dir=both,
        y explicit,
    ] table[x expr=\coordindex, y=pulse_snr_mean_vl_64, y error plus=pulse_snr_sd_vl_64] {\datatable};
    \addlegendentry{VitalLens*};
    \end{axis}
  \end{tikzpicture}
  \caption{Comparing signal-to-noise ratio between VitalLens and VitalLens* on VV-Medium, grouped by participant skin types. VitalLens* was solely trained on PROSIT, while VitalLens was trained on both PROSIT and VV-Africa-Small.}
  \label{fig:impact-skin-type-comparison}
\end{figure}

Next, we want to investigate to what degree the inclusion of VV-Africa-Small in our training data lessened this skin type effect.
For this purpose, we compare VitalLens with \textit{VitalLens*}, a version of VitalLens that is solely trained on the PROSIT test set.
The training set of PROSIT by itself only includes 2 individuals of skin type 6, a number that is massively increased in the combined training data.
Both VitalLens and VitalLens* are then evaluated on VV-Medium.
Figure \ref{fig:impact-skin-type-comparison} compares the results, again grouped by skin type.
It is evident that the addition of VV-Africa-Small led to a sizeable improvement mainly for skin types 5 and 6, but also skin types 3 and 4.

@@
